---
title: "capm_and_apt"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Read in portfolio asset and macroeconomic data:
```{r}
econ = c('CHNGDP','USGDP','EZGDP','US_UNEMP')

shock = c('CHNGDP_Shock','USGDP_Shock','EZGDP_Shock','US_UNEMP_Shock')

finstruments = c('UST_10YR','USFFR','USDRMB','CRUDOIL','CFE_VIX','USDEUR','UST_2YR',
             'SP500_GSCI','USDOIS','UIVE_SP500VALUEETF','USDJPY','USDGBP')

assets = c('VNQ_VANGREALEST','EMB_USDEMRGBOND','LQD_CORPBOND',
            'MUB_MUNIBOND','SHY_1-3USTR','VIG_VANGDIV','IVV_SP500','EEM_MSCIEMERGING',
            'XLE_ENERGYSPDR','EFA_MSCIEAFE','TIP_TIPSBOND')

dataPath = "/Users/chrisolen/Documents/uchicago_courses/deep_learning_and_image_recognition/finance/fin-portfolio-mvo/data"
dat = read.csv(paste(dataPath,"data_cleaned.csv",sep="/"), header=TRUE)
head(dat)

```

### Extract just the portfolio assets, risk free rate, and market indicator:
```{r}
cols.use <- names(dat)[names(dat) %in% assets]
risk.free.rate <- "USFFR"
market.indicator <- "UIVE_SP500VALUEETF"
var.explained = 0.9
full.history <- dat[,c("date",risk.free.rate,market.indicator,cols.use)]
full.history[,"date"] = as.Date(full.history[,"date"], format = "%m/%d/%Y")
head(full.history)
cat("portfolio dimensions:", dim(full.history))

```

### Determine Year(s) of Analysis
```{r}
year <- c(2016,2017)
portfolio <- subset(full.history, format(as.Date(date),"%Y")==year)
head(portfolio)
```

# EF/Capital Asset Pricing Model Approach

### N-Year Mean vs. Variance
```{r}

mean.sd.portfolio <- cbind(sd=apply(portfolio[,-c(1,2)],2,function(z) sd(diff(log(z)),na.rm=TRUE)), mean=apply(portfolio[,-c(1,2)],2,function(z) mean(diff(log(z)),na.rm=TRUE)))
mean.sd.portfolio
```

```{r}
mean.riskfree <- mean(portfolio[,2])/100/250
mean.riskfree
```

```{r}
plot(mean.sd.portfolio[-1,],ylab="Mean",xlab="Sd",pch=19,xlim=c(0,.04),ylim=c(-.0040,.004)) 
points(0,mean.riskfree,col='green',pch=19) 
points(mean.sd.portfolio[1,1],mean.sd.portfolio[1,2],col='red',pch=19) 
text(mean.sd.portfolio[2,1],mean.sd.portfolio[2,2],labels=substring(rownames(mean.sd.portfolio)[2], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[3,1],mean.sd.portfolio[3,2]*-5,labels=substring(rownames(mean.sd.portfolio)[3], 1,3),cex=.8,col="green",pos=2)
text(mean.sd.portfolio[4,1],mean.sd.portfolio[4,2]*5,labels=substring(rownames(mean.sd.portfolio)[4], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[5,1],mean.sd.portfolio[5,2],labels=substring(rownames(mean.sd.portfolio)[5], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[6,1],mean.sd.portfolio[6,2],labels=substring(rownames(mean.sd.portfolio)[6], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[7,1],mean.sd.portfolio[7,2]*-1.5,labels=substring(rownames(mean.sd.portfolio)[7], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[8,1],mean.sd.portfolio[8,2]*6,labels=substring(rownames(mean.sd.portfolio)[8], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[9,1],mean.sd.portfolio[9,2]*6,labels=substring(rownames(mean.sd.portfolio)[9], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[10,1],mean.sd.portfolio[10,2]*6,labels=substring(rownames(mean.sd.portfolio)[10], 1,3),cex=.8,col="blue",pos=2)
text(mean.sd.portfolio[11,1],mean.sd.portfolio[11,2]*6,labels=substring(rownames(mean.sd.portfolio)[11], 1,3),cex=.8,col="blue",pos=2)

```

```{r}
plot(mean.sd.portfolio[-1,],ylab="Mean",xlab="Sd",pch=19,xlim=c(0,.04),ylim=c(-.0040,.004)) 
points(0,mean.riskfree,col='green',pch=19) 
points(mean.sd.portfolio[1,1],mean.sd.portfolio[1,2],col='red',pch=19) 
lines(c(0,3*mean.sd.portfolio[1,1]),c(mean.riskfree,3*mean.sd.portfolio[1,2]),col='red') 
text(mean.sd.portfolio[1,1],mean.sd.portfolio[1,2],labels="market",cex=.8,col="red",pos=2)
text(.0025,mean.riskfree*5,labels="riskfree",cex=.8,col="green",pos=2)

```

### Pull out highest n-year return given level of risk
```{r}
# ef: efficient frontier
bool.ef.asset <- unname(mean.sd.portfolio[,2]/mean.sd.portfolio[,1] == max(mean.sd.portfolio[,2]/mean.sd.portfolio[,1])) 
ratios <- mean.sd.portfolio[,2]/mean.sd.portfolio[,1]
top.ratio <- ratios[bool.ef.asset] 
plot(mean.sd.portfolio[-1,],ylab="Mean",xlab="Sd",pch=19,xlim=c(0,.04),ylim=c(-.0040,.004)) 
points(mean.sd.portfolio[which(bool.ef.asset == TRUE),1],mean.sd.portfolio[which(bool.ef.asset == TRUE),2],col='yellow', pch=19)
points(0,mean.riskfree,col='green',pch=19)
lines(c(0,3*mean.sd.portfolio[which(bool.ef.asset == TRUE),1]),c(mean.riskfree,3*mean.sd.portfolio[which(bool.ef.asset == TRUE),2]),col='orange') 
text(mean.sd.portfolio[which(bool.ef.asset == TRUE),1],mean.sd.portfolio[which(bool.ef.asset == TRUE),2],labels=rownames(mean.sd.portfolio)[which(bool.ef.asset == TRUE)],cex=.8,col="orange",pos=2)
```

### Cumulative Returns of High-Yield/Risk Vs. Least Risky
```{r}
# pull out index of highest return given level of risk
ef.asset.index <- which(bool.ef.asset == TRUE)
ef.asset <- portfolio[,-c(1:2)][ef.asset.index]

# pull out index of least risky
bool.lowest.risk <- unname(mean.sd.portfolio[,1] == min(mean.sd.portfolio[,1]))
lowest.risk.index <- which(bool.lowest.risk == TRUE)
lowest.risk.asset <- portfolio[,-c(1:2)][lowest.risk.index]

# plot
labDates <- paste(format(portfolio[,1],"%m"),format(portfolio[,1],"%d"),sep="/")
matplot(portfolio$date[-1],cbind(cumsum(diff(log(as.matrix(lowest.risk.asset)))),cumsum(diff(log(as.matrix(ef.asset))))),type="l",lty=1,lwd=3,col=c("red","gold"),ylab="Cumulative Returns",xlab="Date",xaxt="n")
axis.Date(1, at = portfolio$date[-1])
legend("topleft", legend=c(colnames(lowest.risk.asset), colnames(ef.asset)), col=c("red", "yellow"), lty=1:2, cex=0.8)

```

### Mixing ef.asset with lowest.risk.asset, to provide the same level of risk as lowest.risk.asset with a higher return
```{r}
# frontier line is line from riskfree rate to ef.asset
frontier.data <- data.frame(rbind(c(mean.riskfree,0), c(mean(diff(log(as.matrix(ef.asset)))), sd(diff(log(as.matrix(ef.asset))))))) 
colnames(frontier.data) <- c("means","sds") 
rownames(frontier.data) <- c("risk.free","ef.asset")

# regressing means of frontier line on risks of frontier line
frontier.line <- lm(means ~ sds, data = frontier.data) 

# extracting lowest.risk.asset risk

portfolio.required.sd <- mean.sd.portfolio[lowest.risk.index,1]
new <- data.frame(sds = portfolio.required.sd)

# predicting the return on the frontier line given EMB-level risk
portfolio.required.mean <- predict(frontier.line, new)

# looking for the point on the frontier line corresponding to the required return
proportion.risky <- (portfolio.required.mean - frontier.data[1,1])/(-frontier.data[1,1] + frontier.data[2,1])
proportion.riskless <- 1-proportion.risky

# create the new portfolio with a mixture of these two assets according to the aboth
mixRFKR.returns <- (portfolio[-1,2]/100/250)*proportion.riskless + diff(log(as.matrix(ef.asset)))*proportion.risky
matplot(portfolio$date[-1], cbind(cumsum(diff(log(as.matrix(lowest.risk.asset)))), cumsum(mixRFKR.returns), cumsum(diff(log(as.matrix(ef.asset))))),type="l",lty=1,lwd=3,col=c("red","purple","gold"), ylab="Cumulative Returns",xlab="Date",xaxt="n")
legend("topleft",legend=c("lowest.risk","mixed", "ef.asset"),col=c("red","purple","gold"),lty=1)
axis.Date(1, at = portfolio$date[-1])

```

### Confirming that the risk of the mixed is the same as that of lowest.risk.asset
```{r}
 c(LowestRisk=sd(diff(log(as.matrix(lowest.risk.asset)))),RiskMixedPortfolio=sd(mixRFKR.returns))
```

### Capital Asset Pricing Model (CAPM)
```{r}
# examining the sensitivity of each individual asset to the market (defined as SP500)
riskfree.daily <- portfolio[,2]/100/365
market <- portfolio[,3]

# we regress the log-difference asset retuns minus the daily FFR against the 'market' returns minus the daily FFR
portfolioBetas <- apply(as.matrix(portfolio[,-c(1,2)]),2,function(z) lm(I(diff(log(z))-riskfree.daily[-length(riskfree.daily)])~-1+I(diff(log(market))-riskfree.daily[-length(riskfree.daily)]))$coefficients)

# betas are estimated as slopes of each excess log returns sample to the SPY excess log returns
plot(1:length(portfolioBetas),portfolioBetas,xlab="Index")
portfolioBetas
```

### Plot the Beta-Mean Returns Space
```{r}
meanReturns <- mean.sd.portfolio[,2]
portfolioBetas[length(portfolioBetas)+1] <- 0
meanReturns[length(meanReturns)+1] <- 0
plot(portfolioBetas, meanReturns, pch = 19)
points(0,0,col='green',pch = 19) 
points(portfolioBetas[1],meanReturns[1],col='red',pch=19) 
text(portfolioBetas[1],meanReturns[1],labels=colnames(portfolio[3]),cex=.8,col="orange",pos=2) 
lines(c(0,4*portfolioBetas[1]),c(0,4*meanReturns[1]),col='red')
```

We recall that CAPM states that those assets that land above the line are undervalued and should be purchased while those below the line are overvalued and should be shorted. Thus, in the former case, we are speculating that the value will go up, and in the latter case, we are speculating that the price will go down. 

### Market-Neutral Portfolio Per CAPM
```{r}
# SML = Security Market Line
SML <- data.frame(a=mean.riskfree, b=mean.sd.portfolio[1,2]-mean.riskfree) 
x <- portfolioBetas[-c(1,length(portfolioBetas))]
y <- SML$a + SML$b * x
```

```{r}
### Long Portfolio ###
# create long portfolio based on SML
longPortfolio <- mean.sd.portfolio[-1,2] > y 
cat("longPortfolio assets:\n",names(mean.sd.portfolio[-1,2][longPortfolio]),"\n")

# plot long portfolio
plot(portfolioBetas[-c(1,length(portfolioBetas))], mean.sd.portfolio[-1,2],ylab="Mean Returns",xlab="Betas",pch=19,xlim=c(0,2))
points(1,mean.sd.portfolio[1,2],col="red",pch=19)
points(0,mean.riskfree,col="green",pch=19) 
lines(c(0,2),c(mean.riskfree,1.75*mean.sd.portfolio[1,2]),col="red") 
points(portfolioBetas[-c(1,length(portfolioBetas))][longPortfolio],mean.sd.portfolio[-1,2][longPortfolio],col="gold",pch=16)

# calculate weights of the long portfolio based on the the distance to the SML
# we assume that the further an asset is above the line, the more undervalued it is, so the more of it we want to be long in
d <- mean.sd.portfolio[-1,2][longPortfolio]-(SML$a + (SML$b*portfolioBetas[-c(1,length(portfolioBetas))][longPortfolio]))
longOnlyWeights <- d/sum(d)
cat("longOnlyWeights:\n",head(longOnlyWeights),"\n")

# calculate the initial value of the weighted long portfolio
longOnlyValue <- as.matrix(portfolio[1,-c(1:3)])[longPortfolio]%*%longOnlyWeights 
cat("initial value of weighted long portfolio:\n",longOnlyValue)
```

We note that assets with negative returns but larger Betas are moving positively with the market and are thus said to be overvalued. 

```{r}
### Short Portfolio ###
cat("short portfolio assets:\n",names(!longPortfolio),"\n")

# calculate weights of the short portfolio based on the the distance to the SML
# we assume that the further an asset is below the line, the more overvalued it is, so the more of it we want to short
d <- (SML$a + (SML$b*portfolioBetas[-c(1,length(portfolioBetas))][!longPortfolio]))-mean.sd.portfolio[-1,2][!longPortfolio]
shortOnlyWeights <- d/sum(d)
cat("short portfolio weights:\n", shortOnlyWeights, "\n")

# calculate the initial value of the weighted short portfolio
shortOnlyValue <- as.matrix(portfolio[1,-c(1:3)])[!longPortfolio]%*%shortOnlyWeights 
cat("short portfolio initial value:", shortOnlyValue, "\n")

```

```{r}
# determine whether long or short portfolio begin at higher value
c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue)
higher = c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue) == max(c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue))

if (higher[1]) {
  portfolioProportion <- longOnlyValue/shortOnlyValue
  cat("Long:Short Proportion ", portfolioProportion, "\n")
  c(longOnlyShares=longOnlyValue/shortOnlyValue, shortOnlyShares=1)

  # value trajectory of the long only portfolio, given the weights
  longValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,longPortfolio]%*%longOnlyWeights 
  # value trajectory of the short only portfolio, given the weights
  shortValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights 
  totallongshortTrajectory.CAPM <- longValueTrajectory%*%portfolioProportion-shortValueTrajectory

  matplot(portfolio$date, totallongshortTrajectory.CAPM,type="l",xlab="year",xaxt="n",ylab="Value of Market-Neutral Portfolio")
  axis.Date(1, at = portfolio$date)
  
} else {
  
  portfolioProportion <- shortOnlyValue/longOnlyValue
  cat("Short:Long Proportion: ", portfolioProportion, "\n")
  c(shortOnlyShares=shortOnlyValue/longOnlyValue, longOnlyShares=1)

  # value trajectory of the long only portfolio, given the weights
  longValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,longPortfolio]%*%longOnlyWeights 
  # value trajectory of the short only portfolio, given the weights
  shortValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights 
  totallongshortTrajectory.CAPM <- longValueTrajectory%*%portfolioProportion-shortValueTrajectory

  matplot(portfolio$date, totallongshortTrajectory.CAPM,type="l",xlab="year",xaxt="n",ylab="Value of Market-Neutral Portfolio")
  axis.Date(1, at = portfolio$date)

}

```

# Arbitrage Pricing Theory Approach

### Convert to log returns:
```{r}
portfolio.returns <- apply(log(portfolio[,-(1:3)]),2,diff)
head(portfolio.returns)
```

### Select PCA factors:
```{r}
n.assets <- dim(portfolio.returns)[2]
portfolio.returns.PCA <- prcomp(portfolio.returns)
summary(portfolio.returns.PCA)$importance[,1:n.assets]

# select number of factors based on accounting for at least var.explained percentage of the variance
prop.var <- summary(portfolio.returns.PCA)$importance[2,1:n.assets]
nFactors <- n.assets - sum(cumsum(prop.var) > var.explained) + 1 
cat("number of factors to be used: ",nFactors)
```

```{r}
# generate factors loadings
factorLoadings <- portfolio.returns.PCA$rotation[,1:nFactors] 
# generate factor scores
factorScores <- portfolio.returns%*%factorLoadings 
# generate zero loading
zeroLoading <- portfolio.returns.PCA$center

# generate approximation of original returns after dim reduction
nFactorAppr <- factorScores%*%t(factorLoadings)

# we want to get a good idea of how well the factor approximations are at predicting the real thing
r.squareds <- mapply(col.ind = 1:ncol(nFactorAppr), function(col.ind){summary(lm(portfolio.returns[,col.ind]~nFactorAppr[,col.ind]))$r.squared})

# get a density plot of all of the r-squared's
plot(density(r.squareds), main = "Distribution of Determination Coefficients") 
# how well factors actually do predict the original assets
abline(v=mean(r.squareds), col = "green") 
# expected amount of variance explained given the number of factors we chose in PCA
abline(v=summary(portfolio.returns.PCA)$importance[3,nFactors],lwd=2,col="red") 
legend("topleft",legend=c("mean r-squared","expected for nFactors"),col=c("green","red" ),lty=1, cex = 0.6)


```

### Use the nFactors PCA components as market factors for APT mode:
```{r}
portfolio.returns.betas <- factorLoadings 

color = grDevices::colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
col=sample(color, n.assets)
matplot(1:nFactors,t(portfolio.returns.betas),type="l",lty=1,xlab="MarketFactors",ylab="Betas",lwd=2, col=col)
legend("bottomleft",legend=rownames(portfolio.returns.betas),lty=1,lwd=2,col=col, cex = .4)

```

The first market factor varies positively with all of the portfolio assets. The second effect has opposite effects on the real estate index than it does on the energy index. 

### Estimation of market prices of risk 
```{r}
# alpha: mean log return for each asset [1, n.assets]
# portfolio.returns.betas: factor loadings for each asset [n.assets, nFactors]
alpha <- zeroLoading
# regressing the average return for each asset on the market risk factors for each asset
mpor.fit <- lm(I(alpha-mean.riskfree)~.-1, data = data.frame(portfolio.returns.betas)) 

# note insignificant factors, if any
summary(mpor.fit)
print("Market Prices of Risk: ", mpor.fit$coefficients)
print("R-Squared: ",  summary(mpor.fit)$r.squared)

# analysis of residuals
modelResiduals <- as.vector(summary(mpor.fit)$residuals) 
hist(modelResiduals)
qqnorm(modelResiduals)
qqline(modelResiduals)
```

### Use the residuals of the equilibrium model to assess the price of each stock relative to the prediction for the time period

```{r}
plot(modelResiduals,type="h",xlab="Stock",ylab="Residual") 
abline(h=0)
 
```

### Create a list of assets recommended for long and short portfolios based on the above:
```{r}
### Long Portfolio ###
longPortfolio.APT <- colnames(portfolio.returns)[modelResiduals > 0] 
dL <- modelResiduals[modelResiduals > 0] 
longPortfolioWeights <- dL/sum(dL) 

# initial avlue of long portfolio
longOnlyValue <- as.matrix(portfolio[1,-(1:3)][longPortfolio.APT])%*%longPortfolioWeights

### Short Portfolio ###
shortPortfolio.APT <- colnames(portfolio.returns)[modelResiduals < 0] 
dS <- modelResiduals[modelResiduals < 0] 
shortPortfolioWeights <- dS/sum(dS)

# initial value of short portfolio
shortOnlyValue <- as.matrix(portfolio[1,-(1:3)][shortPortfolio.APT])%*%shortPortfolioWeights

```

```{r}
# determine whether long or short portfolio begin at higher value
c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue)
higher = c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue) == max(c(longOnlyValue=longOnlyValue, shortOnlyValue=shortOnlyValue))

if (higher[1]) {
  portfolioProportion <- longOnlyValue/shortOnlyValue
  cat("Long:Short Proportion ", portfolioProportion, "\n")
  c(longOnlyShares=longOnlyValue/shortOnlyValue, shortOnlyShares=1)

  # value trajectory of the long only portfolio, given the weights
  longValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,longPortfolio]%*%longOnlyWeights 
  # value trajectory of the short only portfolio, given the weights
  shortValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights 
  totallongshortTrajectory.APT <- longValueTrajectory%*%portfolioProportion-shortValueTrajectory

  matplot(portfolio$date, totallongshortTrajectory.APT,type="l",xlab="year",xaxt="n",ylab="Value of Market-Neutral Portfolio")
  axis.Date(1, at = portfolio$date)
  
} else {
  
  portfolioProportion <- shortOnlyValue/longOnlyValue
  cat("Short:Long Proportion: ", portfolioProportion, "\n")
  c(shortOnlyShares=shortOnlyValue/longOnlyValue, longOnlyShares=1)

  # value trajectory of the long only portfolio, given the weights
  longValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,longPortfolio]%*%longOnlyWeights 
  # value trajectory of the short only portfolio, given the weights
  shortValueTrajectory <- as.matrix(portfolio[,-(1:3)])[,!longPortfolio]%*%shortOnlyWeights 
  totallongshortTrajectory.APT <- longValueTrajectory%*%portfolioProportion-shortValueTrajectory

  matplot(portfolio$date, totallongshortTrajectory.APT,type="l",xlab="year",xaxt="n",ylab="Value of Market-Neutral Portfolio")
  axis.Date(1, at = portfolio$date)

}
```


### lowest.risk.asset vs. ef.asset vs. mixed.risk vs. longshort (CAPM) vs. longshort (APM)
```{r}
matplot(portfolio$date[-1], cbind(cumsum(diff(log(as.matrix(lowest.risk.asset)))), cumsum(mixRFKR.returns), cumsum(diff(log(as.matrix(ef.asset)))), cumsum(diff(log(totallongshortTrajectory.CAPM))), cumsum(diff(log(totallongshortTrajectory.APT)))),type="l",lty=1,lwd=3,col=c("red","purple","gold", "blue","green"), ylab="Cumulative Returns",xlab="Date",xaxt="n")
legend("topleft",legend=c("lowest.risk","mixed.risk", "ef.asset", "longshort.capm", "longshort.apt"),col=c("red","purple","gold", "blue","green"),lty=1)

```



























































