---
title: "pca_risk_scenarios"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ellipse)
library(car)
```

### Read in portfolio asset and macroeconomic data:
```{r}
econ = c('CHNGDP','USGDP','EZGDP','US_UNEMP')

shock = c('CHNGDP_Shock','USGDP_Shock','EZGDP_Shock','US_UNEMP_Shock')

finstruments = c('UST_10YR','USFFR','USDRMB','CRUDOIL','CFE_VIX','USDEUR','UST_2YR',
             'SP500_GSCI','USDOIS','UIVE_SP500VALUEETF','USDJPY','USDGBP')

assets = c('VNQ_VANGREALEST','EMB_USDEMRGBOND','LQD_CORPBOND',
            'MUB_MUNIBOND','SHY_1-3USTR','VIG_VANGDIV','IVV_SP500','EEM_MSCIEMERGING',
            'XLE_ENERGYSPDR','EFA_MSCIEAFE','TIP_TIPSBOND')

dataPath = "/Users/chrisolen/Documents/uchicago_courses/deep_learning_and_image_recognition/finance/fin-portfolio-mvo/data"
dat = read.csv(paste(dataPath,"data_cleaned.csv",sep="/"), header=TRUE)
head(dat)

```

### Extract just the portfolio assets, risk free rate, and market indicator:
```{r}
cols.use <- names(dat)[names(dat) %in% assets]
risk.free.rate <- "USFFR"
market.indicator <- "UIVE_SP500VALUEETF"
var.explained = 0.8
full.history <- dat[,c("date",risk.free.rate,market.indicator,cols.use)]
full.history[,"date"] = as.Date(full.history[,"date"], format = "%m/%d/%Y")
head(full.history)
cat("portfolio dimensions:", dim(full.history))

```

### Determine Year(s) of Analysis
```{r}
year <- c(2016,2017)
portfolio <- subset(full.history, format(as.Date(date),"%Y")==year)
head(portfolio)
```

### Convert to log returns:
```{r}
portfolio.returns <- apply(log(portfolio[,-(1:3)]),2,diff)
head(portfolio.returns)
```

### Select PCA factors:
```{r}
n.assets <- dim(portfolio.returns)[2]
portfolio.returns.PCA <- prcomp(portfolio.returns)
summary(portfolio.returns.PCA)$importance[,1:n.assets]

# select number of factors based on accounting for at least var.explained percentage of the variance
prop.var <- summary(portfolio.returns.PCA)$importance[2,1:n.assets]
nFactors <- n.assets - sum(cumsum(prop.var) > var.explained) + 1 
cat("number of factors to be used: ",nFactors)
```


```{r}
# generate factors loadings
factorLoadings <- portfolio.returns.PCA$rotation[,1:nFactors] 
# generate factor scores
factorScores <- portfolio.returns%*%factorLoadings 
# generate zero loading
zeroLoading <- portfolio.returns.PCA$center

# generate approximation of original returns after dim reduction
nFactorAppr <- factorScores%*%t(factorLoadings)

# we want to get a good idea of how well the factor approximations are at predicting the real thing
r.squareds <- mapply(col.ind = 1:ncol(nFactorAppr), function(col.ind){summary(lm(portfolio.returns[,col.ind]~nFactorAppr[,col.ind]))$r.squared})

# get a density plot of all of the r-squared's
plot(density(r.squareds), main = "Distribution of Determination Coefficients") 
# how well factors actually do predict the original assets
abline(v=mean(r.squareds), col = "green") 
# expected amount of variance explained given the number of factors we chose in PCA
abline(v=summary(portfolio.returns.PCA)$importance[3,nFactors],lwd=2,col="red") 
legend("topleft",legend=c("mean r-squared","expected for nFactors"),col=c("green","red" ),lty=1, cex = 0.6)

# visualize factor loadings plotted against each individual asset's first return of the sample
matplot(1:length(portfolio.returns[1,]),factorLoadings,type="l") 
abline(h=0)

```

In certain cases, factorLoadings may be all positive or all negative across assets. In certain cases, it will depend on the asset for a given factorLoading. 

### Approximation of stock returns for the first day of the time sample
```{r}
approximations.day1 <- factorScores[1,]%*%t(factorLoadings) + zeroLoading 
# plotting approximations for day one versus actual day one
plot(1:length(portfolio.returns[1,]),portfolio.returns[1,]) 
points(1:length(portfolio.returns[1,]),approximations.day1,col="red")

```

### Approximation of stock returns based on first risk factor (PC1)
```{r}
approximations.PC1 <- factorScores[,1]%*%t(factorLoadings[,1]) + zeroLoading 
# plotting approximations for day one versus actual day one
plot(1:length(portfolio.returns),portfolio.returns) 
points(1:length(portfolio.returns),approximations.PC1,col="red")

```

### Plot concentration ellipse for first two factors, denoting 86th and 99th percentiles
```{r}
dtel <- dataEllipse(factorScores[,1:2],levels=c(.99,.86466),xlim=c(-.5,.5),ylim=c(-.3,.5))

# or alternatively:
plot(factorScores[,1:2],xlim=c(-.5,.5),ylim=c(-.3,.5)) 
lines(dtel$`0.99`,col="red") 
lines(dtel$`0.86466`,col="blue")
```


### Create four scenarios corresponding to the extreme points on the 99% concentration ellipse:
```{r}
# west scenario is values corresponding to 1st percentile of first PC
W <- dtel$`0.99`[which.min(dtel$`0.99`[,1]),] 
# east scenario is values corresponding to 99th percentile of the first PC
E <- dtel$`0.99`[which.max(dtel$`0.99`[,1]),] 
# south scenario is values corresponding to 1st percentile of the second PC
S <- dtel$`0.99`[which.min(dtel$`0.99`[,2]),] 
# north scenario is values corresponding to the 99th percentile of the second PC
N <- dtel$`0.99`[which.max(dtel$`0.99`[,2]),] 
(scenarios.99th <- rbind(W,E,N,S))

# plotting on ellipse
plot(factorScores,xlim=c(-.5,.5),ylim=c(-.3,.5)) 
lines(dtel$`0.99`,col="red") 
points(scenarios.99th,pch=16,col="blue")

# calculating and plotting returns: take scenario values and transform back to log returns via factorLoadings
approximationsW <- scenarios.99th["W",]%*%t(factorLoadings[,1:2])+zeroLoading 
approximationsE <- scenarios.99th["E",]%*%t(factorLoadings[,1:2])+zeroLoading 
approximationsN <- scenarios.99th["N",]%*%t(factorLoadings[,1:2])+zeroLoading 
approximationsS <- scenarios.99th["S",]%*%t(factorLoadings[,1:2])+zeroLoading 

# worst case scenario
plot(1:length(approximationsW),approximationsW,col="orange",pch=16,ylim=c(-.05,.05),ylab="Scenario Approximations",xlab="Index") 
# best case
points(1:length(approximationsW),approximationsE,col="magenta",pch=16) 
# second best
points(1:length(approximationsW),approximationsN,col="blue",pch=16) 
# second worst
points(1:length(approximationsW),approximationsS,col="green",pch=16) 
abline(h=0)

rbind(TotalReturns=c(W=sum(approximationsW), E=sum(approximationsE), N=sum(approximationsN), S=sum(approximationsS)),
      AverageReturns=c(W=mean(approximationsW), E=mean(approximationsE), N=mean(approximationsN),S=mean(approximationsS)))

```

### Explore Pearson correlations between assets
```{r}
corr.matrix <- cor(portfolio.returns)
ord <- order(corr.matrix[1,])
xc <- corr.matrix[ord, ord]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc, col=colors[5*xc + 6])

```

### Pearson correlation after removing the influence of the first factor (hedge w.r.t the first factor):
```{r}
corr.matrix.pear <- cor((portfolio.returns-approximations.PC1)) 
ord <- order(corr.matrix.pear[1,])
xc <- corr.matrix.pear[ord, ord]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc, col=colors[5*xc + 6])

```

### Spearman correlation after removing the influence of the first factor (hedge w.r.t the first factor):
```{r}
corr.matrix.spear <- cor((portfolio.returns-approximations.PC1),method="spearman") 
ord <- order(corr.matrix.spear[1,])
xc <- corr.matrix.spear[ord, ord]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc, col=colors[5*xc + 6])

```

### Kendall correlation after removing the influence of the first factor (hedge w.r.t the first factor):
```{r}
corr.matrix.kend <- cor((portfolio.returns-approximations.PC1),method="kendall") # n_comonotonic pairs vs. n_countermonotonic pairs 
ord <- order(corr.matrix.kend[1,])
xc <- corr.matrix.kend[ord, ord]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc, col=colors[5*xc + 6])

```

### Plot Pearson vs. Spearman and Pearson vs. Kendall
```{r}
pear.v.spear <- corr.matrix.pear-corr.matrix.spear
ord1 <- order(pear.v.spear[1,])
xc1 <- pear.v.spear[ord1, ord1]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc1, col=colors[5*xc1 + 6])

pear.v.kend <- corr.matrix.pear-corr.matrix.kend
ord2 <- order(pear.v.kend[1,])
xc2 <- pear.v.kend[ord2, ord2]
colors <- c("#A50F15","#DE2D26","#FB6A4A","#FCAE91","#FEE5D9","white", "#EFF3FF","#BDD7E7","#6BAED6","#3182BD","#08519C")
plotcorr(xc2, col=colors[5*xc2 + 6])



```

Spearman and, in particular, Kendall shows slightly higher correlation than Pearson. 

### Plot empirical copula
```{r}
pairs(apply((portfolio.returns-approximations.PC1)[,1:10],2,rank))
```

Empirical copula show some nonlinear dependency between assets such as VANREAL and ENERGY.