---
title: "risk_metrics"
output: rmarkdown::github_document
---

```{r}
riskScripts <- "/Users/chrisolen/Documents/uchicago_courses/deep_learning_and_image_recognition/finance/fin-portfolio-mvo/risk_scripts"
source(paste(riskScripts,"RMfit.R",sep="/"))
source(paste(riskScripts,"RMeasure.R",sep="/"))
source(paste(riskScripts,"Hill.R",sep="/"))
source(paste(riskScripts,"SimGarcht.R",sep="/"))
library(fGarch)
library(ellipse)

```

### Read in portfolio asset and macroeconomic data:
```{r}
econ = c('CHNGDP','USGDP','EZGDP','US_UNEMP')

shock = c('CHNGDP_Shock','USGDP_Shock','EZGDP_Shock','US_UNEMP_Shock')

finstruments = c('UST_10YR','USFFR','USDRMB','CRUDOIL','CFE_VIX','USDEUR','UST_2YR',
             'SP500_GSCI','USDOIS','UIVE_SP500VALUEETF','USDJPY','USDGBP')

assets = c('VNQ_VANGREALEST','EMB_USDEMRGBOND','LQD_CORPBOND',
            'MUB_MUNIBOND','SHY_1-3USTR','VIG_VANGDIV','IVV_SP500','EEM_MSCIEMERGING',
            'XLE_ENERGYSPDR','EFA_MSCIEAFE','TIP_TIPSBOND')

dataPath = "/Users/chrisolen/Documents/uchicago_courses/deep_learning_and_image_recognition/finance/fin-portfolio-mvo/data"
dat = read.csv(paste(dataPath,"data_cleaned.csv",sep="/"), header=TRUE)
head(dat)

```

### Extract just the portfolio assets, risk free rate, and market indicator:
```{r}
cols.use <- names(dat)[names(dat) %in% assets]
risk.free.rate <- "USFFR"
market.indicator <- "UIVE_SP500VALUEETF"
var.explained = 0.8
nDay.forecastHorizon <- 10
full.history <- dat[,c("date",risk.free.rate,market.indicator,cols.use)]
full.history[,"date"] = as.Date(full.history[,"date"], format = "%m/%d/%Y")
head(full.history)
cat("portfolio dimensions:", dim(full.history))

```

### Determine Year(s) of Analysis
```{r}
year <- c(2016,2017)
portfolio <- subset(full.history, format(as.Date(date),"%Y")==year)
head(portfolio)
```

### Convert to log returns:
```{r}
portfolio.returns <- apply(log(portfolio[,-(1:3)]),2,diff)
head(portfolio.returns)
```

### Risk Metrics

RiskMetrics assumes that the continuously compounded daily rate of return $r_t$ (and therefore the loss variable $x_t$) of a portfolio follows a conditional normal distribution. Such that: 
$$x_t|F_{t-1}~N(0,σ_t^2)$$
…where $σ_t^2$ is the conditional variance of $x_t$ and it evolves over time according to GARCH(1,1):
$$x_t=µ+σ_t^2 ε_t$$
$$σ_t^2=∝σ_{t-1}^2+(1-∝)x_{t-1}^2,1>∝>0$$

Under the GARCH(1,1) model, the conditional variance is proportional to the time horizon k, such that:
$$σ_t^2 [k]=kσ_{t+1}^2$$

VaR given time horizon k and normally distributed loss variable $z$:
$$VaR_{1-p}=Value of Position×z_{1-p}×√k×σ_{t+1} $$

### Calculate risk metrics via Garch and then RMeasure:
```{r}
# fit garch (1,1) with Gaussian innovations to loss variables (xt = -rt)
xt <- -portfolio.returns
gaussianRM.oneDay.95p <- matrix(data=NA,nrow=dim(xt)[2],ncol=2)

for (i in 1:dim(xt)[2]){
  gm <- garchFit(~garch(1,1),data=xt[,i],trace=F) 
  # extract one-day mean and sd from garch
  oneday.mean <- predict(gm,1)[1,1]
  oneday.sd <- predict(gm,1)[1,3]
  rm <- RMeasure(oneday.mean, oneday.sd)
  gaussianRM.oneDay.95p[i,1:2] <- rm$results[1,2:3]
}

colnames(gaussianRM.oneDay.95p) <- c("VaR","ES")
rownames(gaussianRM.oneDay.95p) <- colnames(portfolio.returns)
gaussianRM.oneDay.95p
```

```{r}
# fit garch (1,1) with Student-t innovations to loss variables (xt = -rt)
xt <- -portfolio.returns
stdRM.oneDay.95p <- matrix(data=NA,nrow=dim(xt)[2],ncol=2)

for (i in 1:dim(xt)[2]){
  gm <- garchFit(~garch(1,1),data=xt[,i],trace=F, cond.dist="std")
  # extract one-day mean and sd from garch
  oneday.mean <- predict(gm,1)[1,1]
  oneday.sd <- predict(gm,1)[1,3]
  # extract dof
  dof <- gm@fit$matcoef[5,1]
  rm <- RMeasure(oneday.mean,oneday.sd,cond.dist="std",df=dof)
  stdRM.oneDay.95p[i,1:2] <- rm$results[1,2:3]
}

colnames(stdRM.oneDay.95p) <- c("VaR","ES")
rownames(stdRM.oneDay.95p) <- colnames(portfolio.returns)
stdRM.oneDay.95p
```

### Compare the two one-day risk metrics
```{r}
# Gaussian Garch
gaussianRM.oneDay.95p

# Student-t Garch
stdRM.oneDay.95p
```

### Calculate multi-day risk metrics for garch with Gaussian innovations
```{r}
# fit garch (1,1) with Gaussian innovations to loss variables (xt = -rt)
xt <- -portfolio.returns
gaussianRM.multiDay.95p <- matrix(data=NA,nrow=dim(xt)[2],ncol=2)

for (i in 1:dim(xt)[2]){
  gm <- garchFit(~garch(1,1),data=xt[,i],trace=F)
  # recalling that under the GARCH(1,1) model, the conditional variance is proportional to the time horizon k
  multi.day.predictions <- predict(gm,nDay.forecastHorizon)
  mf <- multi.day.predictions$meanForecast 
  merr <- multi.day.predictions$meanError 
  pmean <- sum(mf) 
  pvar <- sum(merr^2) 
  pstd <- sqrt(pvar) 
  rm <- RMeasure(pmean,pstd)
  
  gaussianRM.multiDay.95p[i,1:2] <- rm$results[1,2:3]
}

colnames(gaussianRM.multiDay.95p) <- c("VaR","ES")
rownames(gaussianRM.multiDay.95p) <- colnames(portfolio.returns)
gaussianRM.multiDay.95p


```

### Calculate multi-day risk metrics for garch with Student-t innovations
```{r}
# We note that the sum of k standardized Student-t distributed random variables is not going to also be distributed as the same standardized Student - unlike in the case of Gaussian innovations. Thus, we need to compute multi-period risk metrics via simulating the garch model:

# fit garch (1,1) with Student-t innovations to loss variables (xt = -rt)
xt <- -portfolio.returns
stdRM.multiDay.95p <- matrix(data=NA,nrow=dim(xt)[2],ncol=2)

for (i in 1:dim(xt)[2]){
  gm <- garchFit(~garch(1,1),data=xt[,i],trace=F, cond.dist="std")
  vol <- volatility(gm)
  alpha1 <- gm@fit$matcoef[3,1]
  omega <- gm@fit$matcoef[2,1]
  alpha1 <- c(omega, alpha1)
  mu <- gm@fit$matcoef[1,1]
  beta1 <- gm@fit$matcoef[4,1]
  dof <- gm@fit$matcoef[5,1]

  # determine the origin of the simulation from the last instance
  orig <- c(xt[length(xt[,i])], vol[length(xt[,i])])
  set.seed(8473625)
  sim <- SimGarcht(h=nDay.forecastHorizon,mu=mu,alpha=alpha1,b1=beta1,df=dof,ini=orig,nter=30000) 
  rr <- sim$rtn

  #VaR
  VaR <- quantile(rr,0.95)
  stdRM.multiDay.95p[i,1] <- VaR
  # ES for p = 0.05 
  idx.p05 <- c(1:30000)[rr > VaR] 
  ES <- mean(rr[idx.p05])
  stdRM.multiDay.95p[i,2] <- ES
  
}

colnames(stdRM.multiDay.95p) <- c("VaR","ES")
rownames(stdRM.multiDay.95p) <- colnames(portfolio.returns)
stdRM.multiDay.95p


```

### Compare two multi-day risk metrics
```{r}
# Gaussian Garch
gaussianRM.multiDay.95p
# Student-t Garch
stdRM.multiDay.95p

```



